---
title: "The Geological Completeness of Paleontological Sampling in North America"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Geological Completeness of Paleontological Sampling in North America}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**Authors:** [Palaeoverse Development Team](https://palaeoverse.org/#about-us)

**Last updated:** `r Sys.Date()`

```{r setup, include = FALSE, eval = TRUE}
knitr::opts_chunk$set(echo = TRUE, include = TRUE, eval = TRUE,
                      out.width = "100%", fig.width = 7.2, fig.height = 6, dpi = 200,
                      fig.path = "")
library(devtools)
load_all()
```

<div style="text-align: justify">

# Introduction

`rmacrostrat` is an R package that allows users to easily retrieve geological data from the [Macrostrat](https://macrostrat.org) database and facilitate analyses within the R environment. This vignette (or tutorial if you prefer) is provided to guide you through the installation process and some of the functionality available within `rmacrostrat`. To do so, we will focus on reconstructing a classical figure from Peters & Heim (2010), which visualizes the changing number, and proportion, of different lithostratigraphic units throughout the Phanerozoic, the last 540 million years.

# Installation

The `rmacrostrat` package can be installed via CRAN or its dedicated [GitHub repository](https://github.com/palaeoverse/rmacrostrat) if the development version is preferred. To install via the CRAN, simply use:

```{r install, eval = FALSE}
install.packages("rmacrostrat")
```

To install the development version, first install the `devtools` package, and then use `install_github` to install `rmacrostrat`` directly from GitHub.

```{r install-dev, eval = FALSE}
install.packages("devtools")
devtools::install_github("palaeoverse/rmacrostrat")
```

You can now load `rmacrostrat` using the default `library` function:

```{r load-library, eval = FALSE}
library(rmacrostrat)
```

**Before we get onto the good stuff, the development team has a small request**. If you use `rmacrostrat`` in your research, please cite the associated publication. This will help us to continue our work in supporting you to do yours. You can access the appropriate citation via:

```{r eval = FALSE}
citation("rmacrostrat")
```

# Context



# Fetching

# Summarisation

# Visualisation

# References

Peters, S.E. and Heim, N.A. 2010. The Geological Completeness of Paleontological Sampling in North America. *Paleobiology* 36(10), pp61--79.

</style>
