---
title: "Plotting a stratigraphic column with rmacrostrat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting a stratigraphic column with rmacrostrat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**Authors:** [Palaeoverse Development Team](https://palaeoverse.org/#about-us)

**Last updated:** `r Sys.Date()`

```{r setup, include = FALSE, eval = TRUE}
knitr::opts_chunk$set(echo = TRUE, include = TRUE, eval = TRUE,
                      out.width = "100%", fig.width = 7.2, fig.height = 6, dpi = 200,
                      fig.path = "")
library(devtools)
load_all()
```

<div style="text-align: justify">

# Introduction

`rmacrostrat` is an R package which allows users to easily retrieve geological data from the [Macrostrat](https://macrostrat.org) database and facilitate analyses within the R environment. This vignette (or tutorial if you prefer) is provided to guide you through the installation process and some of the functionality available within `rmacrostrat`. To do so, we will focus on reproducing a classical figure (i.e., Fig. 14) from [Peters & Heim (2010)](https://www.jstor.org/stable/25609444), which visualizes the changing number--and proportion--of different paleoenvironment lithostratigraphic units throughout the Phanerozoic. Let's get started!

# Installation

The `rmacrostrat` package can be installed via CRAN, or its dedicated [GitHub repository](https://github.com/palaeoverse/rmacrostrat) if the development version is preferred. To install via CRAN, simply use:

```{r install, eval = FALSE}
install.packages("rmacrostrat")
```

To install the development version, first install the `devtools` package, and then use `install_github` to install `rmacrostrat` directly from GitHub.

```{r install-dev, eval = FALSE}
install.packages("devtools")
devtools::install_github("palaeoverse/rmacrostrat")
```

You can now load `rmacrostrat` using the default `library` function:

```{r load-library, eval = TRUE}
library(rmacrostrat)
```

**Before we get onto the good stuff, the development team has a small request**. If you use `rmacrostrat` in your research, please cite the associated publication. This will help us to continue our work in supporting you to do yours. You can access the appropriate citation via:

```{r citation, eval = TRUE}
citation("rmacrostrat")
```

# Context

Quantifying the spatiotemporal distribution of lithostratigraphic units is fundamental to understanding how environments, and their associated biodiversity, have evolved throughout Earth's geological history. Previous work on this theme by [Peters & Heim (2010)](https://www.jstor.org/stable/25609444) focused on *"the geological completeness of paleontological sampling in North America"* with the aim of recognizing and overcoming geologically controlled sampling biases. One element of this article focused on quantifying how the number and proportion of different paleoenvironment lithostratigraphic unit types (e.g., marine, marginal, terrestrial) vary through time. In this vignette, we will revisit this theme and examine how the number and proportion of different lithostratigraphic units--grouped by paleoenvironment type--change throughout the Phanerozoic. To do so, we will make use of the `rmacrostrat` package to fetch data from the [Macrostrat](https://macrostrat.org) database.

# Retrieving data

To quantify how the number and proportion of different lithostraphic units change through time we will focus on the following paleoenvironmental groupings: marine, marginal, and terrestrial. The first step in establishing these groupings is to see what different environments are actually available in [Macrostrat](https://macrostrat.org):

*Hint: Remember that all definitions of data stored in [Macrostrat](https://macrostrat.org) are available via the `def_` suite of functions.*

```{r environments, eval = TRUE}
# What environments are available?
# Hint: Defining no arguments returns all environments
environments <- def_environments()
# See the first few rows
head(environments)
```

From this call, we can see a whole suite of different inferred paleoenvironments are available, which are conveniently grouped into a type (e.g., carbonate, fluvial, lacustrine) and class (i.e., marine and non-marine). Class seems helpful for defining between marine and non-marine environments, but we also want a marginal class, so we need to re-classify some of these environment definitions! 

We can define which paleoenvironments are marginal using the additional information available in the `name` column of `environments`. For this definition, let's consider that any environment that is classified as a delta, beach, barrier island, estuary, lagoon, and tidal flat as a marginal environment. From our available environments (*n* = 87), we identify 22 as marginal environments. This classification is perhaps a little subjective, but for this vignette it provides a good enough representation of marginal environments:

```{r marginal, eval = TRUE}
# Define an object classifying marginal palaeoenvironments
marginal <- c("shallow subtidal", "open shallow subtidal", 
              "lagoonal/restricted shallow subtidal", "paralic indet.", 
              "lagoonal", "coastal indet.", "foreshore", "shoreface",
              "transition zone/lower shoreface", "barrier bar",
              "deltaic indet.", "delta plain", "interdistributary bay",
              "delta front", "prodelta", "marginal marine",
              "deltaic indet.", "delta plain", "lacustrine deltaic indet.",
              "lacustrine delta plain", "lacustrine prodelta", "playa")
```

Now we've classified our marginal paleoenvironments, we can re-classify the `class` column in `environments` so that we have three categories. We can also replace non-marine environments with terrestrial for clarity:

```{r recode, eval = TRUE}
# Update class to be marginal
environments$class[which(environments$name %in% marginal)] <- "marginal"
# Let's also change the name of non-marine class to terrestrial
environments$class[which(environments$class == "non-marine")] <- "terrestrial"
```

Hmmmm our lines of code are getting a little long and hard to read, let's make this simpler and create three `data.frame`s, one for each paleoenvironment.

```{r classes, eval = TRUE}
# Subset dataframe
marine <- subset(environments, class == "marine")
marginal <- subset(environments, class == "marginal")
terrestrial <- subset(environments, class == "terrestrial")
```

Great! So we now have our groupings, but if we want to investigate how paleoenvironment lithostratigraphic units change through time, we need a little more information. Specifically, we need to know the inferred paleoenvironment and age of specific lithostratigraphic units. We can retrieve such information about lithostratigraphic units using the `get_units` function. Conveniently, we can request such data for each paleoenvironment class using the `environ_id` column. 

*Hint: Remember that to retrieve data stored in [Macrostrat](https://macrostrat.org) (e.g., Macrostrat columns, sections, and units), the `get_` suite of functions must be used.*

```{r get_units, eval = TRUE}
# Retrieve marine units
marine <- get_units(environ_id = marine$environ_id)
# Retrieve marginal units
marginal <- get_units(environ_id = marginal$environ_id)
# Retrieve terrestrial units
terrestrial <- get_units(environ_id = terrestrial$environ_id)
# Let's also add the environment class in these data.frames for consistency
marine$environ_class <- "Marine"
marginal$environ_class <- "Marginal"
terrestrial$environ_class <- "Terrestrial"
```

Nice! That was pretty straight forward. However, we notice something weird. Some `unit_id` values in the `marine` dataframe are also present in the marginal and terrestrial dataframes, and vice versa:

```{r duplicate_ids, eval = TRUE}
# How many marine unit IDs are also in terrestrial?
length(which(marine$unit_id %in% terrestrial$unit_id))
```

On investigation, we notice that some lithostratigraphic units are composed of multiple paleoenvironments (various combinations of marine, marginal, and terrestrial). Interesting, but to be expected given transgression and regression cycles! We should treat these differently to our pre-existing groupings as they clump together different paleoenvironments which could capture broader information (e.g., terrestrial and marine organisms). Let's make a new paleoenvironment class called "mixed", and update the environment class accordingly.

```{r create_mixed, eval = TRUE}
# What unit IDs are shared between marine, terrestrial, and marginal?
shared_id <- c(
  # Marine
  marine$unit_id[which(marine$unit_id %in% terrestrial$unit_id)],
  marine$unit_id[which(marine$unit_id %in% marginal$unit_id)],
  # Marginal
  marginal$unit_id[which(marginal$unit_id %in% marine$unit_id)],
  marginal$unit_id[which(marginal$unit_id %in% terrestrial$unit_id)],
  # Terrestrial
  terrestrial$unit_id[which(terrestrial$unit_id %in% marine$unit_id)],
  terrestrial$unit_id[which(terrestrial$unit_id %in% marginal$unit_id)]
)
# Get unique IDs
shared_id <- unique(shared_id)
# Combine dataframes
litho_units <- rbind.data.frame(marine, marginal, terrestrial)
# Recode environment class
litho_units$environ_class[which(litho_units$unit_id %in% shared_id)] <- "Mixed"
# Retain only unique units
litho_units <- unique(litho_units)
```

Great, we can now see that we don't have any duplicate units in our dataset anymore: 

```{r check_units, eval = TRUE}
# Check if any units appear more than once
any(table(litho_units$unit_id) > 1)
```

However, we've just remembered that [Macrostrat](https://macrostrat.org) doesn't just cover North America, and we've pulled units from various continents - D'oh! To make our fairer reproduction of Figure 14 from [Peters & Heim (2010)](https://www.jstor.org/stable/25609444), we ought to focus just on North America. Conveniently, [Macrostrat](https://macrostrat.org) is split into different projects, which tend to cover different regions:

```{r projects, eval = TRUE}
# Check available projects
def_projects()
```

From this call, we can see that North America has a `project_id` of 1. We can use this information to further filter `litho_units`:

```{r n_am, eval = TRUE}
# Filter to N. America
litho_units <- subset(litho_units, project_id == 1)
```

Now we've got our data and done a bit of wrangling, let's get on with summarizing our data.

# Summarizing data

We are interested in how the number and proportion of lithostratigraphic units representing different paleoenvironment classes changes through time. Conveniently, lithostratigraphic units in [Macrostrat](https://macrostrat.org) have age information associated with them, specifically the minimum (`t_age`) and maximum (`b_age`) age in millions of years before present. Using this data, with some handy support functions from the `palaeoverse` R package, we can count the number of lithographic units within each group, and within each time bin, through time. We will use international stratigraphic stage bins as our time bins for this:

```{r counts, eval = TRUE}
# Load palaeoverse R package
library(palaeoverse)
# Generate stage-level time bins
bins <- time_bins(rank = "stage")
# Rename age columns in litho_units to be consistent with our bins
colnames(litho_units)[which(colnames(litho_units) == "b_age")] <- "max_ma"
colnames(litho_units)[which(colnames(litho_units) == "t_age")] <- "min_ma"
# As we are only interested in the Phanerozoic, let's remove any
# data that has a minimum age of more than 541 Ma 
litho_units <- subset(litho_units, min_ma <= 541)
# In case the maximum age of units exceeds 541, but the minimum 
# age is less than 541, let's set max_ma to 541 to still count it 
# if it appears at any point in the Fortunian
litho_units$max_ma[which(litho_units$max_ma > 541)] <- 541
# Bin data using the bin_time function with the "all"
# method. This will bin units into all bins the unit
# overlaps with generating additional rows for each time 
# bin (expect an increase in the number of rows)
litho_units <- bin_time(occdf = litho_units,
                        bins = bins,
                        min_ma = "min_ma", 
                        max_ma = "max_ma",
                        method = "all")
# Calculate the number of environment classes per bin assignment
# Remember, each row is a unique instance of a unit within a bin
counts <- group_apply(occdf = litho_units, 
                      group = c("environ_class", "bin_assignment"),
                      fun = nrow)
# Rename columns to ease reading and merging
colnames(counts) <- c("count", "environment", "bin")
# Merge datasets by bin number
counts <- merge(x = bins, y = counts, by = "bin")
```

# Visualizing data

Now we have our summary of the lithostratigraphic units, we can visualise this data. Let's make two plots, one of the number of units within each paleoenvironment group through time, and another of the proportion of units within each paleoenvironment group through time. To support this data visualization, we will make use of the `ggplot2` (plotting) and `deeptime` (adding a geological timescale axis) R packages.

```{r visualize, eval = TRUE}
# Load data visualization packages
library(ggplot2)
library(deeptime)

# Let's first set our factor levels to replicate the original figure
counts$environment <- factor(counts$environment, levels = c("Marine", "Marginal", 
                                                            "Mixed", "Terrestrial"))

# Generate a plot of raw counts of lithostratigraphic units through time
# grouped by paleoenvironment
ggplot(counts, aes(fill = environment, y = count, x = mid_ma)) + 
  # Stacked bar chart with bar width specified by interval duration
  geom_bar(position = "stack", stat = "identity", width = counts$duration_myr,
           color = "black", linewidth = 0.1) +
  # Label y-axis
  scale_y_continuous("Number of lithostratigraphic units") +
  # Label x-axis and reverse direction
  scale_x_reverse("Time (Ma)") +
  # Data plotting colors
  scale_fill_viridis_d() +
  # Theming 
  theme_bw() +
  theme(legend.title = element_blank(),
        legend.position = "top") +
  # Add geological timescale
  coord_geo()

# Generate a plot of lithostratigraphic units through time grouped by
# paleoenvironment as proportions
ggplot(counts, aes(fill = environment, y = count, x = mid_ma)) + 
  # Proportional bar chart with bar width specified by interval duration
  geom_bar(position = "fill", stat = "identity", width = counts$duration_myr,
           color = "black", linewidth = 0.1) +
  # Label y-axis
  scale_y_continuous("Proportion of lithostratigraphic units") +
  # Label x-axis and reverse direction
  scale_x_reverse("Time (Ma)") +
  # Data plotting colors
  scale_fill_viridis_d() +
  # Theming
  theme_bw() +
  theme(legend.title = element_blank(),
        legend.position = "top") +
  # Add geological timescale
  coord_geo()
```

And that's it! Not too bad, right? There are some clear differences between our generated plots, and those from the original [Peters & Heim (2010)](https://www.jstor.org/stable/25609444) paper (Fig. 14), which is to be expected given that substantially more data is now available in [Macrostrat](https://macrostrat.org), and there may be differences in what we define as "marginal" environments. Despite some differences, some broad-scale patterns are largely consistent, such as the higher proportion of terrestrial lithostratigraphic units in the Mesozoic and Cenozoic, compared to that of the Paleozoic.

Hopefully this vignette has shown you the potential uses for `rmacrostrat` functions and helped provide a workflow for your own analyses. If you have any questions about the package or its functionality, please feel free to join our [Palaeoverse Google group](https://groups.google.com/g/palaeoverse) and leave a question; we'll aim to answer it as soon as possible!

If you're interested in learning more about `rmacrostrat`, don't forget to check out our other vignettes! You can see which ones are available by calling `vignette(package = "rmacrostrat")`.

# References

Gearty, W. 2024. deeptime: Plotting Tools for Anyone Working in Deep Time. R package version 1.1.1, <https://CRAN.R-project.org/package=deeptime>.

Jones, L.A., Gearty, W., Allen, B.J., Eichenseer, K., Dean, C.D., Galván S., Kouvari, M., Godoy, P.L., Nicholl, C.S.C., Dillon, E.M., Flannery-Sutherland, J.T., Chiarenza, A.A. 2022. palaeoverse: A community-driven R package to support palaeobiological analysis. *Methods in Ecology and Evolution*, 14(9), 2205--2215. doi: 10.1111/2041-210X.14099.

Peters, S.E. and Heim, N.A. 2010. The Geological Completeness of Paleontological Sampling in North America. *Paleobiology* 36(10), pp61--79.

Wickham, H. 2016. ggplot2: Elegant Graphics for Data Analysis. *Springer-Verlag New York*.

</div>

```{css echo = FALSE}
p.caption {
  display: none;
}
```
